---
title: "Hands-on Exercise 1: ggplot2 Basics"
author: "Roopa Kaustubha Rayudu"
date: "17 April 2026"
date-format: long
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: cosmo
    highlight-style: github
execute:
  warning: false
  message: false
---

## Overview

This exercise walks through the core building blocks of **ggplot2** — R's grammar-of-graphics plotting library — using exam score data from a cohort of Primary 3 students. We cover everything from basic bar charts to layered plots with statistics and facets.

The data has **7 variables**:

- **Categorical**: `ID`, `CLASS`, `GENDER`, `RACE`
- **Continuous**: `MATHS`, `ENGLISH`, `SCIENCE`

---

## Getting Started

### Loading Packages

We use `pacman` to load `tidyverse` (which bundles ggplot2, readr, and more).

```{r}
pacman::p_load(tidyverse)
```

### Importing the Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
glimpse(exam_data)
```

> **What I see:** 322 students, scores ranging widely — MATHS goes from single digits to near-perfect. Should be interesting to visualise!

---

## Why ggplot2?

Base R can make a histogram in one line:

```{r}
hist(exam_data$MATHS)
```

But ggplot2 gives us a **grammar** — a structured, composable way of building charts. Instead of memorising different function signatures for every plot type, you always think in terms of:

1. **Data** — what am I plotting?
2. **Aesthetics** — what maps to x, y, colour, size?
3. **Geom** — what shape should represent the data?
4. **Extras** — stats, facets, themes, coordinates

---

## The Grammar of Graphics — Layers

ggplot2 is built on Leland Wilkinson's *Grammar of Graphics*. A plot is made of **stacked layers**:

| Layer | What it does |
|---|---|
| `data` | The dataset |
| `aes()` | Maps variables to visual properties |
| `geom_*()` | The actual chart shape |
| `stat_*()` | Statistical transformations |
| `facet_*()` | Split into sub-plots |
| `coord_*()` | Coordinate system |
| `theme_*()` | Non-data styling |

---

## Essential Elements

### Data — the blank canvas

Calling `ggplot()` alone initialises the object with no plot yet — just a blank canvas.

```{r}
ggplot(data = exam_data)
```

### Aesthetics — setting the axes

Adding `aes()` tells ggplot what variable goes where. Here we set the x-axis to MATHS scores.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS))
```

---

## Geometric Objects (geoms)

### Bar Chart — `geom_bar()`

How many students are there per race group?

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar(fill = "#4C72B0", color = "white") +
  labs(title = "Student Count by Race",
       x = "Race", y = "Count") +
  theme_minimal()
```

> **Note:** I added a colour and theme here right away — no reason to leave plots looking dull.

### Dot Plot — `geom_dotplot()`

Each dot = one student's MATHS score. A more granular look at the distribution than a histogram.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_dotplot(binwidth = 2.5,
               dotsize = 0.5,
               fill = "#55A868") +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Distribution of MATHS Scores (Dot Plot)",
       x = "MATHS Score") +
  theme_minimal()
```

> **Why turn off the y-axis?** The raw y-axis in dotplots shows proportions, not counts — it's misleading. Removing it makes the chart cleaner.

### Histogram — `geom_histogram()`

A classic. Default uses 30 bins — usually too many.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 color = "white",
                 fill = "#4C72B0") +
  labs(title = "Distribution of MATHS Scores",
       x = "MATHS Score", y = "Count") +
  theme_minimal()
```

### Histogram by Gender — mapping fill inside `aes()`

When `fill` is inside `aes()`, it maps a variable (GENDER) to colour — ggplot handles the colouring automatically.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           fill = GENDER)) +
  geom_histogram(bins = 20,
                 color = "grey30",
                 alpha = 0.8) +
  scale_fill_manual(values = c("Female" = "#DD8452", "Male" = "#4C72B0")) +
  labs(title = "MATHS Score Distribution by Gender",
       x = "MATHS Score", y = "Count", fill = "Gender") +
  theme_minimal()
```

> **My take:** Males seem to have a wider spread. Density plots (next) will make this clearer.

### Density Plot — `geom_density()`

A smoothed alternative to the histogram, great for comparing distributions.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           colour = GENDER)) +
  geom_density(linewidth = 1) +
  scale_colour_manual(values = c("Female" = "#DD8452", "Male" = "#4C72B0")) +
  labs(title = "MATHS Score Density by Gender",
       x = "MATHS Score", y = "Density", colour = "Gender") +
  theme_minimal()
```

### Boxplot — `geom_boxplot()`

Shows the median, IQR, and outliers for MATHS by gender. Adding `notch = TRUE` helps visually test if medians differ significantly.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS, x = GENDER,
           fill = GENDER)) +
  geom_boxplot(notch = TRUE, alpha = 0.7) +
  scale_fill_manual(values = c("Female" = "#DD8452", "Male" = "#4C72B0")) +
  labs(title = "MATHS Score by Gender (Notched Boxplot)",
       x = "Gender", y = "MATHS Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

> **Notches:** If the notches don't overlap between groups, that's evidence the medians are statistically different.

### Violin Plot — `geom_violin()`

Like a boxplot but shows the full distribution shape — great for spotting bimodal patterns.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS, x = GENDER,
           fill = GENDER)) +
  geom_violin(alpha = 0.7) +
  scale_fill_manual(values = c("Female" = "#DD8452", "Male" = "#4C72B0")) +
  labs(title = "MATHS Score Distribution by Gender (Violin)",
       x = "Gender", y = "MATHS Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Scatter Plot — `geom_point()`

Do students who score well in MATHS also score well in ENGLISH?

```{r}
ggplot(data = exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point(colour = "#4C72B0", alpha = 0.5, size = 1.5) +
  labs(title = "MATHS vs ENGLISH Scores",
       x = "MATHS Score", y = "ENGLISH Score") +
  theme_minimal()
```

### Combining Geoms — Boxplot + Jitter

You can stack multiple geoms. Here the individual data points are overlaid on the boxplot.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS, x = GENDER,
           fill = GENDER)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_point(position = "jitter", size = 0.6, alpha = 0.4) +
  scale_fill_manual(values = c("Female" = "#DD8452", "Male" = "#4C72B0")) +
  labs(title = "MATHS Score by Gender — Boxplot with Points",
       x = "Gender", y = "MATHS Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

> **Why jitter?** Without it, points stack on top of each other and you lose information. Jitter adds a tiny random horizontal offset so you can see density.

---

## Statistics (stat)

### Adding the Mean to a Boxplot

Boxplots show the *median*, not the mean. Here we add the mean as a red diamond.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS, x = GENDER)) +
  geom_boxplot(fill = "#C8D8E8") +
  stat_summary(geom = "point",
               fun = "mean",
               colour = "red",
               size = 4,
               shape = 18) +
  labs(title = "MATHS by Gender — Mean shown as ◆",
       x = "Gender", y = "MATHS Score") +
  theme_minimal()
```

### Best Fit Line on a Scatter Plot

`geom_smooth()` adds a trend line. Using `method = lm` gives a straight linear regression line.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point(colour = "#4C72B0", alpha = 0.4, size = 1.5) +
  geom_smooth(method = lm, linewidth = 0.8,
              colour = "#DD8452", fill = "#DD8452", alpha = 0.15) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  labs(title = "MATHS vs ENGLISH — with Linear Trend",
       x = "MATHS Score", y = "ENGLISH Score") +
  theme_minimal()
```

> **`coord_cartesian()`** forces both axes to the same 0–100 scale so the comparison is fair and not visually distorted.

---

## Facets

Facets split one plot into multiple sub-plots by a category — a powerful way to compare across groups without overplotting.

### `facet_wrap()` — by Class

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 15, fill = "#4C72B0", colour = "white") +
  facet_wrap(~ CLASS) +
  labs(title = "MATHS Score Distribution by Class",
       x = "MATHS Score", y = "Count") +
  theme_minimal()
```

### `facet_grid()` — by Class in a grid

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 15, fill = "#55A868", colour = "white") +
  facet_grid(~ CLASS) +
  labs(title = "MATHS Distribution by Class (Grid Layout)",
       x = "MATHS Score", y = "Count") +
  theme_minimal()
```

> **facet_wrap vs facet_grid:** `facet_wrap` flows panels like text — it wraps into rows automatically. `facet_grid` is rigid and always stays in a single row/column grid.

---

## Coordinates

### Flipping a Bar Chart — `coord_flip()`

When category labels are long, a horizontal bar chart is easier to read.

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar(fill = "#4C72B0") +
  coord_flip() +
  labs(title = "Student Count by Race (Horizontal)",
       x = "Race", y = "Count") +
  theme_minimal()
```

---

## Themes

Themes control the non-data look of a plot. Here's the same chart with three different themes side-by-side in concept (shown individually):

```{r}
base_plot <- ggplot(data = exam_data, aes(x = RACE)) +
  geom_bar(fill = "#4C72B0") +
  coord_flip() +
  labs(x = "Race", y = "Count")

# theme_gray (default)
base_plot + theme_gray() + labs(title = "theme_gray()")
```

```{r}
base_plot + theme_classic() + labs(title = "theme_classic()")
```

```{r}
base_plot + theme_minimal() + labs(title = "theme_minimal()")
```

> **My preference:** `theme_minimal()` — clean, no unnecessary gridlines or grey background. Lets the data speak.

---

## Key Takeaways

By working through this exercise, I've learned:

1. ggplot2 builds plots in **layers** — each `+` adds something new
2. `aes()` *inside* the call maps data to visuals; arguments *outside* set fixed values
3. Geoms determine the chart type — and they can be **combined**
4. `stat_summary()` lets you overlay computed statistics (like means) on any chart
5. Facets are a powerful alternative to colour-coding when you have many groups
6. Themes are purely cosmetic — `theme_minimal()` is clean and professional

---

## References

- Hadley Wickham (2023) [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/)
- Winston Chang (2013) [R Graphics Cookbook](https://r-graphics.org/)
- Kieran Healy (2019) [Data Visualization: A Practical Introduction](https://socviz.co/)
